---
date: 2024-02-27
lesson: PrePro2
topic: Piping / Joins
index: 1
knitr:
  opts_chunk:
    collapse: true
---

# Prepro 2: Demo

```{r}
#| purl: false
#| echo: false
#| output: false

knitr::purl("prepro/Prepro2_Demo.qmd", "prepro/Prepro2_Demo.R", documentation = 0)
```

The code for this demo can also [be downloaded as an R Script](Prepro2_Demo.R) (right click → *Save Target As..*)

## Piping

We want to extract the temperature data from a character string (`temperature`), and then convert the Farenheit value into Celsius according to the following formula, before finally calculating the mean of all the values:

$$°C = (F - 32) \times \frac{5}{9}$$


```{r}
# these temperature are in Kelvin:
temperature <- c("111", "150", "123")


temperature
```


Translated into R-code, this results in the following operation:

```{r}

library("magrittr")

out <- as.numeric(temperature)      # 1. Convert to numeric
out <- subtract(out, 32)            # 2. Subtract 32
out <- multiply_by(out, 5/9)        # 3. Multiply with 5/9
out <- mean(out)                    # 4. Calculate the mean
```

If you want to avoid caching values, this results in a complex, illegible operation.

```{r}
out <- mean(multiply_by(subtract(as.numeric(temperature), 32), 5/9))
```

Piping can be used to keep the code readable and avoid storing intermediate results: *With “piping,” the output of one function becomes the first parameter of the following function.*

```{r}
temperature |>                  
  as.numeric() |>               # 1. Convert to numeric
  subtract(32) |>               # 2. Subtract 32
  multiply_by(5/9) |>           # 3. Multiply with 5/9
  mean()                        # 4. Calculate the mean
```

:::{.callout-important}
- the `|>` pipe operator was first introduced in R `4.1`
- In addition to the *base R* pipe operator, there is also a very similar[^pipe] pipe operator, `%>%`, in the `magrittr` package.
- The <kbd>Ctrl</kbd> +<kbd>Shift</kbd>+<kbd>M</kbd> keyboard shortcut in RStudio inserts a pipe operator.
- By checking the Use native pipe operator setting in RStudio Settings `Tools` → `Global Options` → `Code`, you can control which `pipe operator`, `|>` or `%>%`, is inserted with the above key combination.
:::

[^pipe]: see <https://stackoverflow.com/q/67633022/4139249>

## Joins

```{r}
students <- data.frame(
  Matriculation_No = c(100002, 100003, 200003),
  Student = c("Patrick", "Manuela", "Eva"),
  ZIP = c(8006, 8001, 8820)
)

students

localities <- data.frame(
  ZIP = c(8003, 8006, 8810, 8820),
  LocalityName = c("Zurich", "Zurich", "Horgen", "Wadenswil")
)

localities
```

```{r}
# Load library
library("dplyr")

inner_join(students, localities, by = "ZIP")

left_join(students, localities, by = "ZIP")

right_join(students, localities, by = "ZIP")

full_join(students, localities, by = "ZIP")
```

```{r}
students <- data.frame(
  Matriculation_No = c(100002, 100003, 200003),
  Student = c("Patrick", "Manuela", "Pascal"),
  Residence = c(8006, 8001, 8006)
)

left_join(students, localities, by = c("Residence" = "ZIP"))
```


